Canonical rings of $\bq$-divisors on projective space and Hirzebruch surfaces

### 2015-07-16 surfaces-write-up/surface-write-up.tex

\section{Canonical Rings on Hirzebruch surfaces}
Let $D=\sum_{i=1}^n \alpha_i D_i \in \mathbb{Q} \otimes \di(\hirz_
k).$ such
that $D_1, \ldots, D_4$ are distinct divisors with bi-degrees $(1,0)
, (1,0), (0,1)$, and $(0,1)$ respectively. Let $D_i = V(f_i)$ where
$f_i \in \sco(\bida_i, \bidb_i)$.  Suppose $f_1$ and $f_2$ are independent linear
polynomials in $x_0$ and $x_1$ and $f_3$ and $f_4$ are independent
linear polynomials in $y_0$ and $y_1$.

@@

\begin{proof}
Let $(\bida_i, \bidb_i)$ be the bi-degrees of $f_i$. Notice that if $\gamma
\in (R_D)_d$, then 
\[
	g = \sum_{i = 1}^n {f_i}^{c_i}
\]

\noindent
where each $c_i \ge - \alpha_i c_i$, $\sum_{i=1}^n c_i \bida_i = 0$,
and $\sum_{i=1}^n c_i \bidb_i = 0$. Then we can view $g$ inside of the
lattice $S \subseteq \mathbb{R}^{n + 1}$ given by ordered elements of
the form $(d, c_1, \ldots, c_n)$ satisfying $c_i \ge - d \alpha_i$
for all $i$, $\sum_{i=1}^n c_i \bida_i = 0$, and $\sum_{i = 1}^n c_i \bidb_i =
 0$. From here, determining a spanning set for $(R_D)_d$ reduces to
finding the extremal rays of $S$ by Proposition ~\ref{prop:cone-generation}.

To do this, we extend the method of O'Dorney \cite[Theorem 8]{dorney:canonical}. 
We first view the cone inside of $\mathbb{R}^{n+1}$ given by elements of the
form $(d, c_1, \ldots, c_n)$ satisfying $c_i \ge -d \alpha_i$ for
all $i$ and $\sum_{i = 1}^n c_i (\bida_i + \bidb_i) = 0$. By O'Dorney, this
has extremal rays given by

\[
	e_i = (1, -\alpha_1, \ldots, -\alpha_{i-1}, \frac{\sum_{j \ne i}
	\alpha_j (\bida_j + \bidb_j)}{\bida_i + \bidb_i}, -\alpha_{i + 1},
	\ldots, -\alpha_m).
\]

Let $\Delta^{n-1}$ be the simplex with vertices $e_1, \ldots e_n$.
We can intersect $\Delta^{n-1}$ with the hyperplane $H$ given by
$\sum_{i=1}^n \bida_i X_i = 0$  to get a subspace $S \subseteq \Delta
^{n-1}$. The extremal rays of $S$ must lie on the edges of $\Delta
^{n - 1}$ \todo{Peter: Do we need to cite/argue this, or is it clear?}.
Thus, the extremal rays of $S$ are given by the set of $e_i$
contained in $H$ and by the points $e_{i, j}$ that are the
intersections of $H$ with edges of the simplex connecting the
distinct $e_i, e_j$ not contained in $\in H$.
\todo{more detail above}

@@

Let $E$ be the set of $\epsilon_i$ and $\epsilon_{i,j}$ given 
above, satisfying the respective conditions.
The lattice points corresponding to elements of $E$ satisfies the conditions of Proposition ~\ref{prop:cone-generation}, so by applying that proposition, $R_D$ is generated in 
degree less than
\[
	\sigma = \sum_{i\in \Te(D)} \gcd(\bida_i, \bidb_i)\ell_i + \sum_{\substack{
	i \in \Tp(D) \\	j \in \Tm(D)}} (\bida_i \bidb_j- \bida_j \bidb_i)\ell_{i,j}.
\]
\end{proof}

Let $x_1, \ldots x_r$ be the generators of $R_D$ in degree less than $\sigma$ (as given by Lemma ~\ref{lem:hirz-generators}), and let $\phi$ be the surjection $k[x_1, \ldots x_r] \to R_D$.  We can factor $\phi$ through 
\[
	\Sigma =  \langle u^d z_1^{c_1} \cdots z_n^{c_n} : c_i \geq -d
	\alpha_i, \sum_{i=1}^{n} \bida_i c_i = \sum_{i=1}^{n} \bidb_i c_i.
	\rangle 
\]
into
\[
\begin{tikzcd}
	\bk[x_1,\ldots, x_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
	x_i \ar r & u^{d_i}z_1^{c_{i1}} \cdots z_n^{c_{in}} \ar r & u^{d_i}f_1^{c_{i1}} \cdots f_n^{c_{in}}.
\end{tikzcd}
\]

Applying Lemma ~\ref{lem:bound-ker-chi}, we can bound the degrees of generation of $\ker(\chi)$ below
$2 \sigma$.
Finally, we calculate the degree of generation of $\psi$:

@@

To find such a $\beta$, we simply write $f_i$ as a 
polynomial in $f_1, \ldots f_4$. Furthermore, these
generate all relations, since they allow us to reduce any $\prod_{j =
1}^n f_j^{r_j}$ to the canonical form (see Proposition ~\ref{prop:cone-generation}).  \todo{Peter: I think the canonical form given in the proposition isn't the one we actually want.}

Note that we can view the relations from Equation
~\ref{eqn:hirz-relations-psi} \todo{is this the right ref?} as an element of $S$ given
by $(d, c_1, \ldots, c_{i-1}, c_i + 1, c_{i+1}, \ldots , c_n)$.  Let $S'\subseteq (\mathbb{R}_{\ge 0})^{n+1}$ be the
set of $(d, c_1, \ldots, c_{i-1}, c_i + 1, c_{i+1}, \ldots , c_n)$ satisfying $c_i \ge -\alpha_i d$,
$\bida_j + \sum_{i = 1}^m \bida_i c_i = 0$, and $\bidb_j + \sum_{i=1}^n \bidb_i c_i
= 0$.  A careful computation deduces that $\gamma (d, c_1, \ldots, c_{i-1}, c_i + 1, c_{i+1}, \ldots, c_n)$ is of the form 
\[
	E + \sum_{k\in \Te} r_k e_k + \sum_{\substack{h \in \Tp \\ k \in \Tm}} r_{h,k} e_{h,k}
\]

\noindent
where $E$ is $\mathbb{R}_{\ge 0}$
combination of elements in 
\[
	\{e_i \mbox{ if } i \in S\} \cup \{e_{i, j}: j \in \Tm \mbox{ if }
	i	\in \Tp \} \cup \{e_{j, i}: j \in \Tp \mbox{ if } i \in \Tm \}.
\]

>>>>

\section{Canonical Rings on Hirzebruch surfaces}
Let $D=\sum_{i=1}^n \alpha_i D_i \in \mathbb{Q} \otimes \di(\hirz_
m).$ such
that $D_1, \ldots, D_4$ are distinct divisors with bi-degrees $(1,0)
, (1,0), (0,1)$, and $(0,1)$ respectively. Let $D_i = V(f_i)$ where
$f_i \in \sco(\bida_i, \bidb_i)$.  
Then, $f_1$ and $f_2$ are independent linear
polynomials in $x_0$ and $x_1$ and $f_3$ and $f_4$ are independent
linear polynomials in $y_0$ and $y_1$. \todo{Talk a little about Hirzebruch surfaces before this paragraph; especially, describe what it's sections look like.}

@@

\begin{proof}
%Let $(\bida_i, \bidb_i)$ be the bi-degrees of $f_i$. 
Notice that if $\gamma
\in (R_D)_d$, then 
\[
	g = \sum_{i = 1}^n {f_i}^{c_i}
\]

\noindent
where each $c_i \ge - \alpha_i c_i$, $\sum_{i=1}^n c_i \bida_i = 0$,
and $\sum_{i=1}^n c_i \bidb_i = 0$. Then we can any view $g$ inside the
lattice $S \subseteq \mathbb{R}^{n + 1}$ given by ordered elements of
the form $(d, c_1, \ldots, c_n)$ satisfying $c_i \ge - d \alpha_i$
for all $i$, $\sum_{i=1}^n c_i \bida_i = 0$, and $\sum_{i = 1}^n c_i \bidb_i =
 0$. From here, determining a spanning set for $(R_D)_d$ reduces to
finding the extremal rays of $S$ by Proposition ~\ref{prop:cone-generation}.

To do this, we extend the method of O'Dorney \cite[Theorem 8]{dorney:canonical}. 
We first consider the cone inside of $\mathbb{R}^{n+1}$ given by elements of the
form $(d, r_1, \ldots, r_n)$ satisfying $r_i \ge -d \alpha_i$ for
all $i$ and $\sum_{i = 1}^n r_i (\bida_i + \bidb_i) = 0$.  This
has extremal rays given by

\[
	e_i := (1, -\alpha_1, \ldots, -\alpha_{i-1}, \frac{\sum_{j \ne i}
	\alpha_j (\bida_j + \bidb_j)}{\bida_i + \bidb_i}, -\alpha_{i + 1},
	\ldots, -\alpha_m).
\]

Let $\Delta^{n-1}$ be the simplex with vertices $e_1, \ldots e_n$.
We can intersect $\Delta^{n-1}$ with the hyperplane $H$ given by
$\sum_{i=1}^n \bida_i X_i = 0$  to get a subspace $S \subseteq \Delta
^{n-1}$. The extremal rays of $S$ must lie on the edges of $\Delta
^{n - 1}$ \todo{Peter: Do we need to cite/argue this, or is it clear?}.
Thus, the extremal rays of $S$ are given by the $e_i$'s
contained in $H$ and the intersection points $e_{i, j}$ of $H$
and the edges $\overline{e_i e_j}$ of the simplex between
distinct $e_i, e_j$ not contained in $H$.
\todo{more detail above}

@@

The set of $e_i$ and $e_{i,j}$ given above satisfies the conditions of Proposition ~\ref{prop:cone-generation}, so by applying that proposition, $R_D$ is generated in 
degrees less than
\[
	\sigma = \sum_{i\in \Te(D)} \gcd(\bida_i, \bidb_i)\ell_i + \sum_{\substack{
	i \in \Tp(D) \\	j \in \Tm(D)}} (\bida_i \bidb_j- \bida_j \bidb_i)\ell_{i,j}.
\]
\end{proof}

Let $x_1, \ldots x_r$ be the generators of $R_D$ in degrees less than $\sigma$ (as given by Lemma ~\ref{lem:hirz-generators}), and let $\phi$ be the surjection $k[x_1, \ldots x_r] \to R_D$.  We can factor $\phi$ through 
\[
	\Sigma =  \langle u^d z_1^{c_1} \cdots z_n^{c_n} : c_i \geq -d
	\alpha_i, \sum_{i=1}^{n} \bida_i c_i = \sum_{i=1}^{n} \bidb_i c_i.
	\rangle 
\]
into
\[
\begin{tikzcd}
	\bk[x_1,\ldots, x_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
	x_i \ar r & u^{d_i}z_1^{c_{i1}} \cdots z_n^{c_{in}} \ar r & u^{d_i}f_1^{c_{i1}} \cdots f_n^{c_{in}}.
\end{tikzcd}
\]

Applying Lemma ~\ref{lem:bound-ker-chi}, we can bound the degree of generation of $\ker(\chi)$ below
$2 \sigma$.
Finally, we calculate the degree of generation of $\psi$:

@@

To find such a $\beta$, we simply write $f_i$ as a 
polynomial in $f_1, \ldots f_4$. Furthermore, these
generate all relations, since they allow us to reduce any $\prod_{j =
1}^n f_j^{r_j}$ to the canonical form where $r_i = - \lfloor d \alpha_i \rfloor$ whenever $i > 4$.

%Note that we can view a relation from Equation
%~\ref{eqn:hirz-relations-psi} as an element of $S$ given
%by $(d, c_1, \ldots, c_{i-1}, c_i + 1, c_{i+1}, \ldots , c_n)$.  
Let $S'\subseteq S$ be the
set of $(d, c_1, \ldots, c_{i-1}, c_i + 1, c_{i+1}, \ldots , c_n)$ satisfying $c_i \ge -\alpha_i d$,
$\bida_i + \sum_{j = 1}^m \bida_j c_j = 0$, and $\bidb_i + \sum_{j=1}^n \bidb_j c_j
= 0$.  A careful computation deduces that the lattice point corresponding to $y_i \prod_{j=1}^n {y_j}^{c_j}$ is of the form 
\[
	E + \sum_{k\in \Te} r_k e_k + \sum_{\substack{h \in \Tp \\ k \in \Tm}} r_{h,k} e_{h,k}
\]

\noindent
where $E$ is in degree $\mathbb{R}_{\ge 0}$
combination of elements in 
\[
	\{e_j: j\in \{i\} \cap S\} \cup \{e_{i, j}: j \in \Tm \mbox{ if }
	i	\in \Tp \} \cup \{e_{j, i}: j \in \Tp \mbox{ if } i \in \Tm \}.
\]
